---
title: "B Allele Frequencies"
author: "Jacob Carey, Steven Cristiano, and Robert Scharpf"
date: \today
output: BiocStyle::pdf_document
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{B Allele Frequencies}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc} 
---

# Introduction

start with simulating data even if unrealistic. help get models that works in
ideal situation. RS and SC discussed - would like to have an interpretation of
the copy number, the components themselves. the BAF will help to some degree,
at least with the bigger ones. some other thigns that SC worked on that will be
help. BAF is interesting as application of MixtureModel. Nice QC for HMM. any
big CNP you find w/ HMM, nice way to verify it. This can be rare. This will be
best as a standalone vignette and another app of MixtureModel. show how it ties
in QC analyses. At the end of main vignette, talk about interpreting copy
number. maing thing prev done was look at variances of components and
separation of components. looked at starting value. if first component mean is
< -1, it means that first component is copy number 0 (homozygous deletion). for
some CNPs is it diploid? deletion? copy number 2, 3 for purposes of vignette,
what we can use is data is HapMap data? potentially a lot of work. Extract what
we need to do 10? RS is looking into what HapMap data is available.

```{r lib} 
library(CNPBayes)
```

Here we simulate batch data with 10 batches but only 1 component. As can be
seen in the plot, there is a batch effect. If batches were ignored, a
MarginalModel would likely find 2 components, when there is in truth only 1.

# Simulation
```{r lrr-sim}
obs <- c(250, 231, 329, 250, 187, 328, 250, 250, 500, 288)
obs.label <- character(0)

for (batch in 1:length(obs)) {
   obs.label <- c(obs.label, rep(letters[batch], obs[batch])) 
}

means <- c(0.03856340, -0.17588675, -0.02816801, -0.08564261, -0.21721350,
           0.12363354, -0.13442788, 0.01568428, 0.07118810, 0.17810444)

vars <- c(0.005558076, 0.008895078, 0.009313795, 0.008954961, 0.007649359,
          0.008733442, 0.010344712, 0.007078906, 0.006794232, 0.007428676)

means <- as.matrix(means)
vars <- as.matrix(vars)

sim.data <- simulateBatchData(N=sum(obs),
                              batch=obs.label,
                              theta=means,
                              sds=sqrt(vars),
                              p=1)

plot <- plot(sim.data)
```

Next we simulate possible B Allele Frequency data corresonding to the above log
R ratio data. 

```{r baf-sim}
baf <- c(rnorm(floor(sum(obs) / 3), 0, 0.06),
         rnorm(floor(sum(obs) / 3), 0.5, 0.06),
         rnorm(floor(sum(obs) / 3), 1, 0.06))
baf[baf < 0] <- 0
baf[baf > 1] <- 1

hist(baf, breaks=100)
```

Finally, we fit mixture models of varying component sizes to the B Allele
Frequency data and use `bic` to identify the best fit.

```{r fit}
model1 <- MarginalModel(baf, k=1)
model2 <- MarginalModel(baf, k=2)
model3 <- MarginalModel(baf, k=3)

model1 <- posteriorSimulation(model1)
model2 <- posteriorSimulation(model2)
model3 <- posteriorSimulation(model3)

bic(model1)
bic(model2)
bic(model3)
```

As is seen, `bic` correctly selects the three component model.
