---
title: "Inspecting Convergence of a Model"
author: "Jacob Carey, Steven Cristiano, and Robert Scharpf"
date: "`r doc_date()`"
output: BiocStyle::pdf_document
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Overview of CNPBayes package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc} 
---

# Introduction

A Markov Chain Monte Carlo posterior simulation should be visually inspected to assess convergence.

```{r lib}
# load CNPBayes
suppressMessages(library(CNPBayes))

# load packages for manipulating and visualizing data
suppressMessages(library(dplyr))
suppressMessages(library(tidyr))
suppressMessages(library(ggplot2))
```

# Workflow

The number of starting values, burnin MCMC simulatioms, and MCMC simulations
after burnin are controlled by a `McmcParams` object. Here, we specify a small
number so that the example runs quickly.

```{r post-sim}
set.seed(1)
N <- 200
n <- 81
lrr <- replicate(N, mean(rnorm(n)))
mp <- McmcParams(iter=100, burnin=00, nStarts=10)
sb.list <- MarginalModelList(data=lrr, mcmc.params=mp, k=1:4)
## ignore the warnings -- label switching occurs because model is overfit
sb.list <- posteriorSimulation(sb.list)
```

One approach for selecting the best model is to compute the marginal likelihood
for each of the models and select the model with the largest marginal
likelihood. Note, the marginal likelihood is only accurate if the models have
converged. Additional burnin iterations and iterations following burnin may be
needed. For chains with a high autocorrelation, specifying the thin parameter
will increase the independence of the simulations saved to the chain.

```{r marginal-lik, eval=FALSE}
m.lik <- marginalLikelihood(sb.list)
m.lik
```

```{r plot1}
 data1 <- as.data.frame(theta(chains(sb.list[[1]]))) %>%
    mutate(iter=1:100) %>%
    gather(component, theta, V1) %>%
    mutate(model=1)

data2 <- as.data.frame(theta(chains(sb.list[[2]]))) %>%
    mutate(iter=1:100) %>%
    gather(component, theta, V1:V2) %>%
    mutate(model=2)

data3 <- as.data.frame(theta(chains(sb.list[[3]]))) %>%
    mutate(iter=1:100) %>%
    gather(component, theta, V1:V3) %>%
    mutate(model=3)

data4 <- as.data.frame(theta(chains(sb.list[[4]]))) %>%
    mutate(iter=1:100) %>%
    gather(component, theta, V1:V4) %>%
    mutate(model=4)

data <- bind_rows(data1, data2, data3, data4) %>%
    mutate(component=gsub("V", "", component))

ggplot(data, aes(x=iter, y=theta)) +
    geom_line(aes(colour=component, linetype=component)) +
    facet_wrap(~model, nrow=2, ncol=2) +
    theme_classic() +
    xlab("")
```

```{r plot2}
data_2.4 <- data %>%
    filter(model > 1, (iter <= 20 | iter >= 80)) %>%
    mutate(iter.cat=cut(iter, c(0, 20, 79, 100),
                        c("bottom", "middle", "top"))) %>%
    group_by(component, iter.cat, model) %>%
    mutate(cat.model.median=median(theta)) %>%
    ungroup() %>%
    mutate(box.cat=paste(iter.cat, component))

ggplot(data_2.4, aes(x=box.cat, y=theta)) +
    geom_boxplot(aes(colour=box.cat)) +
    geom_hline(yintercept=0.0, linetype="dashed", colour="gray") +
    facet_wrap(~model, scales="free_x") +
    guides(colour=FALSE) +
  theme_classic() +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
    xlab("")
```


A shortcut for plotting the chains (here, for the k=3 model):

```{r chains}
plist <- ggSingleBatchChains(sb.list[[3]])
## chains of component-specific parameters
plist[["comp"]]
## single-value parameters
plist[["single"]]
```
# References
