---
title: "Overview of CNPBayes package"
author: "Jacob Carey, Steven Cristiano, and Robert Scharpf"
date: \today
output: BiocStyle::pdf_document
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Overview of CNPBayes package}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc} 
---

# Introduction

CNPBayes models multi-modal densities via a hierarchical Bayesian
Gaussian mixture model.  The major application of this model is the
estimation of copy number at copy number polymorphic loci (CNPs). Two
versions of the mixture model are implemented.  A *standard* model,
referred to as a *marginal* model, that has one mean and standard
deviation for each component, and a *batch* model with batch-specific
means and standard deviations. Approximation of the posterior is by
Markov Chain Monte Carlo (MCMC) written in C++ using the Rcpp package
[@Rcpp].  

For an EM-implementation of Gaussian mixture models for CNPs, see the
Bioconductor package CNVtools [@Barnes2008].  A Bayesian extension of
this model by some of the same authors was developed to automate the
analysis of the Welcome Trust Case Control Consortium (WTCCC) genotype
data [@cardin] and implemented in the R package CNVCALL
(http://niallcardin.com/CNVCALL).

This vignette provides a concise workflow for fitting mixture models
in large array-based genome-wide association studies.  We refer the
reader to other vignettes included with this package for details
regarding implementation.

```{r lib} 
library(CNPBayes)
```

# Workflow

## Delineate CNPs for each ancestry group

```{r find_cnps}
se <- readRDS(system.file("extdata", "simulated_se.rds", package="CNPBayes"))
grl <- readRDS(system.file("extdata", "grl_deletions.rds", package="CNPBayes"))
```

## Fitting mixture models at each CNP

 Independently for each CNP, we group the samples into batches using
the `collapseBatch` function.

Some Example code

For array-based estimates and germline genomes, there are typically
between 1 and 4 copy number states at any given CNP.  Since we do not
know the number of components a priori, we fit a model for each k.
In addition, we fit models with and without a term for batch.

## Selecting a model

Marginal likelihood is estimated for each of the 8 models and the
Bayes' factors is used to select one of these models for further
study.


## Extracting MAP estimates and posterior probabilities

How to do this for the selected model

## QC

Visualize the results.  Check BAFs.

# References
